### 2025-08-14 chapters/benchmark.tex
revised 3.4 to include cache configurations

\begin{table}[H]
\centering
\scriptsize
\caption{Median CPU Cycles for Kyber Operations}
\begin{tabular}{|l|r|r|}
\hline
\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} \\
\hline
indcpa\_keypair & 1,870,200 & 1,702,800 \\
\hline
indcpa\_enc     & 38,376,800 & 2,174,400 \\
\hline
indcpa\_dec     & 39,060,800 & 618,400 \\
\hline
kyber\_keypair  & 1,870,200 & 1,800,300 \\
\hline
kyber\_encaps   & 38,376,800 & 2,285,900 \\
\hline
kyber\_decaps   & 39,060,800 & 2,921,800 \\
\hline
\end{tabular}
\label{tab:kyber_median_cycles}
\end{table}

@@

@@

@@

%\begin{figure}[h]
%    \centering
%    \includegraphics[width=0.7\textwidth]{figures/throughput_comparison.pdf}
%    \caption{Benchmark throughput under native and enclave execution}
%    \label{fig:throughput}
%\end{figure}

@@

@@

\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}

The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications. Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.

Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution. This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.

As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design. Memory usage during enclave execution remained stable, with no indications of paging or contention. System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.

From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed. This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.

>>>>

\begin{table}[h]
\centering
\scriptsize
\begin{tabular}{|l|r|r|p{4.5cm}|}
\hline
\textbf{Operation} & \textbf{Parallel Run10 Median (cycles)} & \textbf{Sequential Run10 Median (cycles)} & \textbf{Rationale / Notes} \\
\hline
indcpa\_keypair & 1,870,200 & 1,702,800 & Baseline IND-CPA key generation cost for setting up public/private pairs. \\
\hline
indcpa\_enc     & 38,376,800 & 2,174,400 & Parallel vs sequential encryption overhead comparison. \\
\hline
indcpa\_dec     & 39,060,800 & 618,400   & Decryption efficiency gap between modes. \\
\hline
kyber\_keypair  & 1,870,200 & 1,800,300 & Full Kyber keypair generation timing. \\
\hline
kyber\_encaps   & 38,376,800 & 2,285,900 & Encapsulation performance with shared secret derivation. \\
\hline
kyber\_decaps   & 39,060,800 & 2,921,800 & Decapsulation latency and handshake completion time. \\
\hline
\end{tabular}
\caption{Median CPU cycles for Kyber operations with concise context to fit page width.}
\label{tab:kyber_median_cycles}
\end{table}

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.75\linewidth]{figures/enclave_memory_pagefaults.png}
%   \caption{Resident Set Size (RSS) memory usage and page fault counts during Kyber enclave runs, showing stable memory consumption and minimal paging overhead.}
%   \label{fig:memory_pagefaults}
% \end{figure}

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.75\linewidth]{figures/context_switches_comparison.png}
%   \caption{Involuntary context switches during native versus enclave execution, highlighting a nearly 200$\times$ increase under enclave protection.}
%   \label{fig:context_switches}
% \end{figure}

@@

@@

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.75\linewidth]{figures/enclave_parallelism_stability.png}
%   \caption{Effect of parallel enclave instances on execution stability. Failures increase sharply beyond 5 concurrent runs due to PMP region limitations.}
%   \label{fig:parallelism_stability}
% \end{figure}

@@

\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}
\label{sec:workload-sensitivity}

This section presents a comparative analysis of enclave-induced overheads across three fundamentally different workload types: cryptographic (Kyber), general-purpose embedded (CoreMark), and integer-heavy synthetic benchmarks (Dhrystone). The goal is to understand how diverse computational patterns react to the constraints and runtime characteristics of trusted enclave execution, particularly under a constrained hardware configuration.

All evaluations were performed on a minimal RISC-V platform comprising a single in-order core and 64~MB of RAM, consistent with prior sections. Each workload was executed both natively and within the secure enclave environment, under identical operating conditions. Enclave runs were repeated ten times sequentially to reduce interference, and measurements were recorded using Linux time utilities and cycle counters embedded in the enclave runtime. System-level statistics, such as context switches, resident set size (RSS), and page fault counts, were also gathered to provide a holistic view of the overheads.

Among the three workloads, Kyber proved to be the most sensitive to enclave-related performance impacts. The median number of CPU cycles for its \texttt{kyber\_decaps} operation increased from approximately 2.9 million in the native case to over 3.5 million when executed inside the enclave—a roughly 1.2$\times$ increase. However, this difference was magnified in terms of wall-clock time: the native execution completed in approximately 10.7 seconds, whereas the enclave version averaged 29.2 seconds, representing a 2.7$\times$ slowdown. This discrepancy suggests that memory encryption and integrity verification add nontrivial latency per memory access, disproportionately affecting workloads with dense memory usage or pointer-heavy data structures, which are characteristic of lattice-based cryptographic primitives like Kyber.

In stark contrast, CoreMark exhibited the least performance degradation under enclave execution. Wall-clock time increased only slightly, from a native average of 101.5 seconds to 107.3 seconds in the enclave, corresponding to a 1.06$\times$ overhead. This minimal slowdown is attributed to CoreMark’s balanced workload structure, combining arithmetic operations, control logic, and memory accesses in a manner that is less sensitive to cryptographic memory protection. CoreMark’s throughput declined modestly, from a native range of 967--988 iterations per second to an enclave-protected range of 919--939, indicating that while overheads exist, they are amortized over the longer duration and lower frequency of secure memory operations.

Dhrystone occupied a middle ground between Kyber and CoreMark in terms of sensitivity. The wall-clock time for Dhrystone increased from 15.3 seconds natively to 24.7 seconds under enclave protection, yielding an overhead factor of approximately 1.6$\times$. Despite its primary reliance on integer arithmetic and control structures, Dhrystone is sensitive to memory system performance due to its short execution time and tight loops. The secure memory mechanisms in the enclave added consistent latency, resulting in degraded throughput (395,000--408,000 Dhrystone operations per second) when compared to the native baseline of approximately 400,000 DPS. These results confirm that even relatively simple benchmarks experience measurable slowdown under enclave protection, though to a lesser extent than cryptographic workloads.

Across all three benchmarks, system-level metrics revealed a consistent pattern in enclave execution behavior. In particular, the number of involuntary context switches per run consistently exceeded 66,000 for each workload. This value is nearly two orders of magnitude greater than in native runs, where fewer than 400 involuntary context switches were recorded. The uniformity of this value across distinct workloads indicates that the overhead is primarily systemic, driven by the enclave runtime itself rather than specific application behavior. These context switches are likely associated with secure page handling, timer interrupts, and asynchronous enclave runtime maintenance operations. Despite their frequency, the impact appears bounded and does not significantly impair CPU-bound computation, as confirmed by the preservation of user-mode CPU time dominance in all executions.

Memory behavior across enclave runs remained remarkably stable and comparable to native execution. All workloads reported a resident set size (RSS) close to 10~KB, with no major page faults and at most one minor fault per run. This result indicates that the enclave’s memory allocator and secure paging subsystem are sufficiently robust to handle working sets without inducing swapping or paging penalties, even under constrained RAM conditions.

These observations are summarized in Table~\ref{tab:sensitivity-overheads}, which compares native and enclave wall-clock times, overhead factors, and context switch counts across all three benchmarks. Figure~\ref{fig:workload_overhead} visually contrasts the wall-clock overhead imposed by enclave execution, highlighting Kyber’s pronounced sensitivity and CoreMark’s relative resilience.

\begin{table}[h]
\centering
\begin{tabular}{|l|c|c|c|p{3.5cm}|}
\hline
\textbf{Benchmark} & \textbf{Native Time (s)} & \textbf{Enclave Time (s)} & \textbf{Overhead Factor} & \textbf{Involuntary CS} \\
\hline
Kyber     & 10.7   & 29.2   & 2.73$\times$  & 66,000 \\
\hline
Dhrystone & 15.3   & 24.7   & 1.61$\times$  & 66,000 \\
\hline
CoreMark  & 101.5  & 107.3  & 1.06$\times$  & 66,000 \\
\hline
\end{tabular}
\caption{Enclave Execution Sensitivity Across Workloads. Overhead factors indicate slowdown relative to native execution.}
\label{tab:sensitivity-overheads}
\end{table}

% \begin{figure}[htbp]
%   \centering
%   \includegraphics[width=0.7\linewidth]{figures/workload_overhead_comparison.png}
%   \caption{Relative increase in wall-clock execution time under enclave protection across three workload types.}
%   \label{fig:workload_overhead}
% \end{figure}

In summary, this workload sensitivity analysis reveals a clear hierarchy of enclave overhead impacts based on workload characteristics. Cryptographic workloads, particularly those like Kyber with memory-intensive operations and frequent data-dependent access patterns, are disproportionately affected by the secure memory fabric’s latency and the enclave runtime’s background operations. In contrast, workloads with predictable access patterns or longer execution durations, such as CoreMark, exhibit substantially lower performance degradation. These insights are critical for system designers seeking to deploy trusted execution environments in resource-constrained systems, as they inform trade-offs between security, performance, and workload type.

### 2025-08-14 chapters/system-architecture.tex
revised 3.4 to include cache configurations

\begin{table}[h]
\centering
\begin{tabular}{|l|c|c|c|p{7.5cm}|}
\hline
\textbf{Config Name} & \textbf{RAM} & \textbf{Cores} & \textbf{Execution} & \textbf{Rationale} \\
\hline
Low-End Baseline     & 64M         & 1             & Sequential         & Serves as a minimal configuration, enabling the establishment of baseline performance under constrained conditions. Useful for measuring the raw impact of enclave mechanisms in resource-starved environments. \\
\hline
Balanced Parallel    & 128M        & 2             & Parallel           & Captures early-stage scaling behavior with modest multicore setups. Helps isolate the overhead introduced by enclave entry/exit in light parallel workloads. \\
\hline
Mid-Range Sequential & 256M        & 4             & Sequential         & Provides a view into how increasing core count and memory capacity affect performance in the absence of thread-level parallelism. Useful for understanding system scalability independent of synchronization effects. \\
\hline
High-End Parallel    & 512M        & 6             & Parallel           & Represents a robust configuration where memory and cores are sufficiently provisioned. Designed to observe near-ideal parallel execution and potential scalability limits of the enclave framework. \\
\hline
Max Capacity Stress  & 2G          & 10            & Parallel           & Pushes the simulated platform to its upper bounds. Intended to explore saturation points, resource contention, and diminishing returns under heavy system load. \\
\hline
\end{tabular}
\caption{Summary of representative configurations used in performance evaluation.}
\label{tab:configurations}
\end{table}

In preliminary testing, a number of alternative configurations were explored to validate the diversity and coverage of the selected set. These included scenarios such as single-core systems with excessively large memory allocations and many-core setups with highly constrained memory. However, these combinations typically resulted in either underutilized resources or pathological behavior (e.g., excessive paging or context switching), without contributing new insight beyond what was captured by the five chosen profiles. For example, increasing memory size on a single-core system had negligible impact due to CPU-bound bottlenecks, while increasing core counts in systems with limited memory often led to high levels of involuntary context switching and scheduling overhead.

In addition to varying core and memory resources, three distinct cache configurations were tested across all system profiles to assess the sensitivity of enclave and native workloads to cache architecture. These configurations included:

\begin{itemize}
    \item A baseline setup featuring 8\,KB, 4-way set-associative caches with an LRU replacement policy
    \item An enlarged cache with 32\,KB size and 8-way associativity, reflecting more generous cache resources
    \item A fixed-size cache using FIFO and random eviction policies to model alternative eviction behaviors
\end{itemize}

All configurations used a consistent block size of 64 bytes to preserve spatial locality across workloads. Although the cache modeling plugin captures system-wide statistics and does not differentiate between enclave and non-enclave activity, observed trends—such as shifts in cache miss rates and access patterns—were analyzed in conjunction with benchmark-level behavior to infer their contribution to observed performance variations.

To enable direct and consistent comparisons, each benchmark was executed in both native and enclave modes under every configuration. This mirrored design allowed for clear isolation of the performance penalties introduced by enclave execution and helped disentangle architectural effects from security-related overhead.

>>>>

\begin{table}[h]
\centering
\footnotesize
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|l|c|c|c|c|c|c|l|}
\hline
\textbf{Configuration} & \textbf{RAM} & \textbf{Cores} & \textbf{Exec.} &
\textbf{D‑Cache} & \textbf{I‑Cache} & \textbf{Policy} & \textbf{Assoc./Blk (D/I)} \\
\hline
Low‑End Baseline &
64\,MB & 1 & Seq. &
4\,KB & 4\,KB & LRU & 2‑way / 32B,\ 2‑way / 32B \\
\hline
Balanced Parallel &
128\,MB & 2 & Par. &
8\,KB & 8\,KB & FIFO & 4‑way / 64B,\ 4‑way / 64B \\
\hline
Mid‑Range Sequential &
256\,MB & 4 & Seq. &
16\,KB & 8\,KB & Random & 8‑way / 32B,\ 4‑way / 64B \\
\hline
High‑End Parallel &
512\,MB & 6 & Par. &
8\,KB & 16\,KB & LRU & 2‑way / 64B,\ 8‑way / 32B \\
\hline
Max Capacity Stress &
2\,GB & 10 & Par. &
32\,KB & 32\,KB & FIFO & 16‑way / 64B,\ 16‑way / 64B \\
\hline
\end{tabular}
\caption{Summary of representative configurations used in performance evaluation.}
\label{tab:configurations}
\end{table}

In preliminary testing, a wider range of system specifications was explored to ensure that the final selection of five configurations (Table~\ref{tab:configurations}) provided both diversity and coverage. Extreme cases—such as single‑core systems with disproportionately large memory or many‑core systems with severely limited memory—were excluded, as they tended to either underutilize resources or exhibit pathological behaviour (e.g., excessive paging or context‑switching) without yielding additional insight. For example, adding memory to a single‑core system had negligible impact due to CPU‑bound bottlenecks, whereas increasing core counts with insufficient memory frequently caused scheduler overhead and contention.

In addition to varying memory capacity, core count, and execution mode (sequential or parallel), each configuration was paired with a distinct cache arrangement to assess the sensitivity of enclave and native workloads to cache architecture. The cache parameters are summarised in Table~\ref{tab:configurations} and include:

\begin{itemize}
    \item Small caches (4--8\,KB) with low associativity and an LRU replacement policy
    \item Mid‑sized caches (8--16\,KB) with mixed associativity and FIFO or random replacement
    \item Large caches (32\,KB) with high associativity and FIFO eviction
\end{itemize}

All caches used fixed block sizes of 32 or 64\,B, as listed in the table, to maintain consistent spatial locality across workloads. While the cache modelling plugin measures aggregate system statistics without distinguishing enclave from non‑enclave activity, observed trends in miss rates and access patterns were correlated with benchmark behaviour to help attribute performance differences.

For each of the five configurations, all benchmarks were executed in both native and enclave modes. This mirrored arrangement facilitated a clear isolation of enclave‑specific overheads from general architectural effects.
